---
title: "Storm Data Analysis 1950-2011"
author: "Susan Lively Klug"
date: "August 20, 2015"
output: html_document
---
# Storm Data Analysis

### Answering the questions
1) Across the United States, which types of events are most harmful with respect to population health?
2) Across the United States, which types of events have the greatest economic consequences?

## Data Processing

Original data and documentation locations as of 8/20/2015

https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf

Download and read into the raw_data variable.
Use cache for this chunk.
```{r, cache=TRUE}

if (!file.exists("StormData.csv.bz2")) {
     download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                   "./StormData.csv.bz2")
}

raw_data <- read.csv("StormData.csv.bz2")

```
Investigate choices for defining "most harmful with respect to population health"
```{r}
library(dplyr)
names(raw_data)
dim(raw_data)
type <- unique(raw_data$EVTYPE)
type_df <- data.frame(type)
dim(type_df)
```
That is a lot of EVTYPEs, 985.  
There are some EVTYPEs with Summaries.  These will overlap with the data
that they are summarizing, so we filter them out.

```{r}
tdf <- filter(type_df, !grepl('Summary', type))
dim(tdf)
```
921 different types is still going to make an unreadable plot, but we only care
about the 'most harmful' so we can do some elimination of lower ranges once we
select the interesting columns.

Except: some of the EVTYPE are mangled and typo issues that should be combined.

"Harmful to the Population Health" in this analysis refers to Fatalities and
Injuries to persons.  The property or economic damage will be in the second
section of the analysis.

From names(raw_data) we know that we are interested in columns "INJURIES" 
and "FATALITIES" and the columns "EVTYPE"

```{r}
columns.of.interest = c("EVTYPE", "INJURIES", "FATALITIES")
pop.health <- select(raw_data, one_of(columns.of.interest))
summary(pop.health)
```

We are safe to remove any rows that are zero for both INJURIES and FATALITIES
as they will not help us in figuring out the "Maximum Damage"
instead of subsetting one could assign them all to the "no damage" category

```{r}
sum( ( pop.health$INJURIES == 0 ) & ( pop.health$FATALITIES == 0) )
pop.health.sm <- filter(pop.health, 
             !(( pop.health$INJURIES == 0 ) & ( pop.health$FATALITIES == 0)))
```
Still doesn't remove many of the EVTYPEs.
Some of the EVTYPES are mangled versions of each other and should be 
combined.  We can fix the lower case using toupper from dplyr.
There are numerous spelling errors and versions of the same EVTYPE
that need to be brute force changed to match.

```{r}
library(reshape2)
dfn = as.data.frame(sapply(pop.health.sm, toupper))

## S3 method for class 'data.frame'
newdfn <- melt(dfn, id.vars=c("EVTYPE") )

# Need to make them allcaps then more will match with group_by
# In dfn rows with matching EVTYPE and Variable
# should be summed and collapsed.
newdfn$value <- as.integer(newdfn$value)
df <- newdfn %>% group_by(EVTYPE, variable) %>% summarise_each(funs(sum))
df <- df[!(df$value == 0 ), ]
```

Now down to some gritty substituting to get to the valid EVTYPES
gsub(pattern, replacement, x)

```{r}
df$EVTYPE = gsub("AVALANCE", "AVALANCHE", df$EVTYPE)
df$EVTYPE = gsub("WINDS", "WIND", df$EVTYPE)
df$EVTYPE = gsub("LANDSLIDES", "LANDSLIDE", df$EVTYPE)
df$EVTYPE = gsub("SEAS", "SEA", df$EVTYPE)
df$EVTYPE = gsub("ROADS", "ROAD", df$EVTYPE)
df$EVTYPE = gsub("CURRENTS", "CURRENT", df$EVTYPE)

df$EVTYPE = gsub("COASTAL FLOOD.*", "COASTAL FLOOD", df$EVTYPE)
df$EVTYPE = gsub("COASTALSTORM", "COASTAL STORM", df$EVTYPE)
df$EVTYPE = gsub(".*FLASH FLOOD.*", "FLASH FLOOD", df$EVTYPE)
df$EVTYPE = gsub("FLOODING", "FLOOD", df$EVTYPE)
df$EVTYPE = gsub("FLOOD.*", "FLOOD", df$EVTYPE)
df$EVTYPE = gsub("MINOR FLOOD", "FLOOD", df$EVTYPE)
df$EVTYPE = gsub("TIDAL FLOOD", "FLOOD", df$EVTYPE)
df$EVTYPE = gsub("URBAN.* FLOOD", "FLOOD", df$EVTYPE)
df$EVTYPE = gsub("RIVER FLOOD", "FLOOD", df$EVTYPE)
df$EVTYPE = gsub("HIGH WATER", "FLOOD", df$EVTYPE)
df$EVTYPE = gsub("RAPIDLY.*.WATER", "FLOOD", df$EVTYPE)
df$EVTYPE = gsub(".*FLD", "FLOOD", df$EVTYPE)

df$EVTYPE = gsub(".*FIRE.*", "WILDFIRE", df$EVTYPE)
df$EVTYPE = gsub("WILD FIRE", "WILDFIRE", df$EVTYPE)

df$EVTYPE = gsub("RAINS", "RAIN", df$EVTYPE)
df$EVTYPE = gsub("RAINFALL", "RAIN", df$EVTYPE)
df$EVTYPE = gsub("EXCESSIVE RAIN", "HEAVY RAIN", df$EVTYPE)
df$EVTYPE = gsub("TORRENTIAL RAIN", "HEAVY RAIN", df$EVTYPE)

df$EVTYPE = gsub("WINTER WEATHER.*", "WINTER WEATHER", df$EVTYPE)
df$EVTYPE = gsub("RAIN/SNOW", "EXTREME COLD/WIND CHILL", df$EVTYPE)
df$EVTYPE = gsub("GLAZE/ICE STORM", "WINTER STORM", df$EVTYPE)
df$EVTYPE = gsub(".*EXTREME.*CHILL", "EX WIND CHILL", df$EVTYPE)
df$EVTYPE = gsub("EXTREME.*CHILL", "EX WIND CHILL", df$EVTYPE)
df$EVTYPE = gsub("COLD.*", "COLD", df$EVTYPE)
df$EVTYPE = gsub("FOG AND COLD", "FREEZING FOG", df$EVTYPE)
df$EVTYPE = gsub("UNS.*.COLD", "COLD", df$EVTYPE)
df$EVTYPE = gsub("EXT.*.COLD", "EXTREME COLD/WIND CHILL", df$EVTYPE)
df$EVTYPE = gsub("HIGH.*.COLD", "EXTREME COLD/WIND CHILL", df$EVTYPE)
df$EVTYPE = gsub("RECORD.*.COLD", "EXTREME COLD/WIND CHILL", df$EVTYPE)
df$EVTYPE = gsub("SNOW.*.COLD", "EXTREME COLD/WIND CHILL", df$EVTYPE)
df$EVTYPE = gsub("BLACK ICE", "FROST/FREEZE", df$EVTYPE)
df$EVTYPE = gsub("FREEZING DRIZZLE", "FROST/FREEZE", df$EVTYPE)
df$EVTYPE = gsub("FREEZING SPRAY", "FROST/FREEZE", df$EVTYPE)
df$EVTYPE = gsub("FREEZING RAIN", "FROST/FREEZE", df$EVTYPE)
df$EVTYPE = gsub("FREEZE", "FROST/FREEZE", df$EVTYPE)
df$EVTYPE = gsub("FROST", "FROST/FREEZE", df$EVTYPE)
df$EVTYPE = gsub("FROST.*.FREEZE", "FROST/FREEZE", df$EVTYPE)
df$EVTYPE = gsub("ICE STORM", "ICE", df$EVTYPE)
df$EVTYPE = gsub(".*ICE.*", "ICE STORM", df$EVTYPE)
df$EVTYPE = gsub("WINTRY.*", "WINTER WEATHER", df$EVTYPE)
df$EVTYPE = gsub("SNOW.*", "SNOW", df$EVTYPE)
df$EVTYPE = gsub("^SNOW.*", "WINTER STORM", df$EVTYPE)
df$EVTYPE = gsub("LIGHT SNOW.*", "WINTER STORM", df$EVTYPE)
df$EVTYPE = gsub("EX*.SNOW.*", "HEAVY STORM", df$EVTYPE)
df$EVTYPE = gsub("THUNDERSNOW", "HEAVY STORM", df$EVTYPE)
df$EVTYPE = gsub("BLOWING SNOW", "HEAVY STORM", df$EVTYPE)
df$EVTYPE = gsub("EXCESSIVHEAVY STORM", "WINTER STORM", df$EVTYPE)
df$EVTYPE = gsub(".*HAIL.*", "HAIL", df$EVTYPE)
df$EVTYPE = gsub("WINTER STORM.*", "WINTER STORM", df$EVTYPE)
df$EVTYPE = gsub(".*ICY.*", "WINTER WEATHER", df$EVTYPE)
df$EVTYPE = gsub("EX WIND CHILL", "EXTREME COLD/WIND CHILL", df$EVTYPE)

df$EVTYPE = gsub(".*DROUGHT.*", "DROUGHT", df$EVTYPE)
df$EVTYPE = gsub(".*EX.*HEAT", "EXCESSIVE HEAT", df$EVTYPE)
df$EVTYPE = gsub(".*HEAT W.*", "HEAT", df$EVTYPE)
df$EVTYPE = gsub("RECORD HEAT", "HEAT", df$EVTYPE)
df$EVTYPE = gsub(".*WARM.*", "HEAT", df$EVTYPE)

df$EVTYPE = gsub(".*FOG", "DENSE FOG", df$EVTYPE)

df$EVTYPE = gsub(".*RIP CURRENT.*", "RIP CURRENT", df$EVTYPE)
df$EVTYPE = gsub(".*SURF.*", "HIGH SURF", df$EVTYPE)
df$EVTYPE = gsub("STORM SURGE.*", "STORM SURGE", df$EVTYPE)

df$EVTYPE = gsub("WATERSPOUT.*", "WATERSPOUT", df$EVTYPE)
df$EVTYPE = gsub("HIGH WIND.*", "HIGH WIND", df$EVTYPE)
df$EVTYPE = gsub(".*LIGHT.*", "LIGHTNING", df$EVTYPE)
df$EVTYPE = gsub("STORMS", "STORM", df$EVTYPE)
df$EVTYPE = gsub("WIND STORM", "WIND", df$EVTYPE)
df$EVTYPE = gsub("RAIN/WIND", "STRONG WIND", df$EVTYPE)
df$EVTYPE = gsub("TROPICAL STORM.*", "TROPICAL STORM", df$EVTYPE)
df$EVTYPE = gsub("THUNDERST.*", "THUNDERSTORM", df$EVTYPE)
df$EVTYPE = gsub("TSTM", "THUNDERSTORM", df$EVTYPE)
df$EVTYPE = gsub(".*HURRI.*", "HURRICANE", df$EVTYPE)
df$EVTYPE = gsub("WIND.*", "WIND", df$EVTYPE)
df$EVTYPE = gsub(".*TORN.*", "TORNADO", df$EVTYPE)
df$EVTYPE = gsub("MARINE THUNDERSTORM", "MARINE THUNDERSTORM WIND", df$EVTYPE)
df$EVTYPE = gsub("^THUNDERSTORM$", "THUNDERSTORM WIND", df$EVTYPE)
df$EVTYPE = gsub("^THUNDERTORM.*", "THUNDERSTORM WIND", df$EVTYPE)
df$EVTYPE = gsub("WIND WIND", "WIND", df$EVTYPE)
df$EVTYPE = gsub("NON THUNDERSTORM WIND", "HIGH WIND", df$EVTYPE)
df$EVTYPE = gsub("^WIND$", "HIGH WIND", df$EVTYPE)
```

At this point visual inspection shows Tornado at an order of magnitude 
more damaging than any other event.  We might be able to drop the '1's

```{r}
m <- median(df$value)
df <- filter(df, (value > m ) )

# followed by more group and summarise
df <- df %>% group_by(EVTYPE, variable) %>% summarise_each(funs(sum))

unique(df$EVTYPE)

```

I would most like to see a bar chart with EVTYPE along the Y value.
This was a quick way to be able to read the thirty plus types of events.
But to narrow things down to executive summary let us look at the top ten.

```{r}
library(ggplot2)
pop.health <- head(df[order(df$value, decreasing = T), ])
pop.health
ggplot(pop.health, aes(x=EVTYPE, y=value, fill=variable)) +
     geom_bar(stat="identity", position=position_dodge())+
     ggtitle("Population Harm by Event") +
     coord_flip()

```
## Greatest Economic Consequences
```{r}
columns.of.interest = c("EVTYPE", "PROPDMG", "PROPDMGEXP")
econ.prop <- select(raw_data, one_of(columns.of.interest))
summary(econ.prop)
columns.of.interest = c("EVTYPE", "CROPDMG","CROPDMGEXP")
econ.crop <- select(raw_data, one_of(columns.of.interest))
summary(econ.crop)
```
Reviewing the values of expenses related to Property and Crops.
PROPDMG is the numeric value of the damage, PROPDMGEXP is the magnitude where 'B'=billions
and the other factors are much smaller.

We can safely view only the 'B' cases for crop and property damage.

```{r}
econ.prop <- filter(econ.prop, PROPDMGEXP=='B')
econ.prop <- select(econ.prop, EVTYPE, PROPDMG)
econ.prop <- econ.prop %>% group_by(EVTYPE) %>% summarise_each(funs(sum))
econ.prop$EVTYPE = gsub(".*HURRI.*", "HURRICANE", econ.prop$EVTYPE)
econ.prop$EVTYPE = gsub(".*TORN.*", "TORNADO", econ.prop$EVTYPE)
econ.prop$EVTYPE = gsub(".*FIRE.*", "WILDFIRE", econ.prop$EVTYPE)

econ.crop <- filter(econ.crop, CROPDMGEXP=='B')
econ.crop <- select(econ.crop, EVTYPE, CROPDMG)
econ.crop <- econ.crop %>% group_by(EVTYPE) %>% summarise_each(funs(sum))

e <- full_join(econ.crop, econ.prop, by="EVTYPE")
e[is.na(e)] <- 0
e$damages <- e$CROPDMG + e$PROPDMG
e <- arrange(e, desc(damages))
e <- transform(e, EVTYPE = reorder(EVTYPE, damages))
ggplot(e, aes(x=EVTYPE, y=damages)) +
     geom_bar(stat="identity") +
     ggtitle("Economic Damages by Weather Event") +
     coord_flip() +
     labs(x = "Billions of dollars", y = "Event type")
```

## Results

